---
title: "Recurrence Quantification Analysis"
output:
  html_notebook: default
---

# Data Exploration
This dataset originates from the traffic signal system, published via the Victoria (Australia) open transport data platform, and contains traffic volumes recorded every 15 minutes throughout March 2026 for multiple intersections. For this group project, we use the combined approach, where detector values are aggregated into a single total traffic volume per site at each timestamp. Detectors are individual sensors at an intersection, typically assigned to specific lanes or traffic movements, that count how many vehicles pass during each 15-minute interval. By summing these detector readings, we obtain a clean and consistent site level time series for analyzing temporal patterns and comparing traffic flow across sites.
```{r}
library(readr)
library(here)

library(ggplot2)

library(nonlinearTseries)

library(tseriesChaos)

library(crqa)
```


```{r}
traffic_data <- read_csv(here("data", "traffic_detectors_combined_100_161_164.csv"))
```

Below is a preview of the dataset. Each row represents a 15-minute interval at a specific site, where traffic volumes from all detectors have been aggregated into a single site-level measurement. The key variables are `site_total_volume`, which provides the traffic count and serves as the main time series for analysis, `timestamp`, which ensures the correct temporal order and regular spacing of observations, and `site_no`, which allows comparison of traffic dynamics across different intersections.

```{r}
head(traffic_data)
```


```{r}
ggplot(traffic_data, aes(x = timestamp, y = site_total_volume)) +
  geom_line(color = "steelblue") +
  facet_wrap(~ site_no, ncol = 1, labeller = labeller(site_no = function(x) paste("Site Number =", x))) +
  labs(
    title = "Traffic Volume Over Time by Site",
    x = "Time",
    y = "Traffic Volume"
  )
```

```{r}
ggplot(traffic_data, aes(x = factor(site_no), y = site_total_volume)) +
  geom_boxplot(fill = "steelblue") +
  labs(
    title = "Distribution of Traffic Volume by Site",
    x = "Site Number",
    y = "Traffic Volume"
  ) 
```

```{r}
by(traffic_data$site_total_volume, traffic_data$site_no, summary)
```


# RQA
## RQA - Single Site Analysis

Based on the data exploration, site 164 was selected for further analysis. This site exhibits the highest variability, with an interquartile range from 185.8 to 1139.2, as well as the highest median traffic volume (Mdn = 777.5), compared to site 161 (Mdn = 548.0) and site 100 (Mdn = 278.0). Furthermore, the observed temporal data exhibit a regularity, which is expected to result in meaningful recurrence patterns.

```{r}
# Select purely site 164 data
site164 <- traffic_data[traffic_data$site_no == 164, ]

# They should be ordered by just in case
site164 <- site164[order(site164$timestamp), ]

# Extract volumes
x <- site164$site_total_volume

```


```{r}

x_scaled <- scale(x)[, 1]

ami <- mutualInformation(x_scaled, lag.max = 96, do.plot = TRUE)
delay <- 29

print(data.frame(lag = 0:96, ami = ami$mutual.information))
```
To estimate the delay parameter τ for RQA, Average Mutual Information (AMI) was computed for the scaled site 164 time series across lags 0 to 96. A maximum lag of 96 was chosen because each lag unit corresponds to one 15-minute interval, meaning 96 lags represents exactly one full day (96 × 15 = 1440 minutes). This ensures the AMI search spans a complete daily cycle. 

The AMI curve dropped sharply after lag 0 and did not show an early local minimum. This pattern is likely caused by the difference between very low night-time traffic volumes and high daytime peaks, which makes smaller within day patterns less visible in the mutual information plot. The first local minimum occurred at lag 29 (AMI = 0.378), corresponding to approximately 7.25 hr. This time interval matches the approximate separation between the morning and evening traffic peaks, which are the two main recurring traffic states at site 164.


```{r}
# Embedding (m) 
fnn <- false.nearest(x_scaled, m = 20, d = 29, t = 1, eps = sd(x_scaled))
plot(fnn)
print(fnn)

embed <- 5
```
The embedding dimension, M, was estimated using the False Nearest Neighbors (FNN) method. This method identifies how many delayed versions of the time series are needed to unfold the underlying system dynamics. The FNN fraction was calculated for embedding dimensions from 1 to 20, using the previously selected delay of τ = 29. The results showed a steep decrease from M = 1 (FNN = 44.0%) to M = 2 (FNN = 13.7%), followed by a slower decline at higher dimensions. After M = 5, where the FNN fraction reached 3.3%, further increases in embedding dimension produced only small improvements. Therefore, M = 5 was selected as a embedding dimension because it captures the main dynamics of the traffic system without adding unnecessary complexity.

```{r}
# Radius (r)

radius_candidates <- seq(0.1, 2.0, by = 0.05)
rr_values <- numeric(length(radius_candidates))

for (i in seq_along(radius_candidates)) {
  result <- crqa(
    ts1         = x_scaled,
    ts2         = x_scaled,
    delay       = 29,
    embed       = 5,
    radius      = radius_candidates[i]
  )
  rr_values[i] <- result$RR
}

plot(radius_candidates, rr_values, type = "b", xlab = "Radius", ylab = "Recurrence Rate (%)", main = "Recurrence Rate by Radius")
abline(h = c(2, 5), col = "red", lty = 2)

print(data.frame(radius = radius_candidates, RR = rr_values))

```
The radius parameter was selected by inspecting recurrence rates across a range of candidate radius values from 0.10 to 2.00, computed in steps of 0.05. For each candidate radius, a recurrence quantification analysis was run using the previously determined parameters of τ = 29 and M = 5, and the resulting recurrence rate was recorded. 

Following the lecture guideline that continuous behavioral signals should ideally have a recurrence rate between 2% and 5%, with more deterministic signals placed closer to the lower end of this range, a radius of 0.55 was chosen. This radius produced a recurrence rate of 2.02%, indicating that the traffic volume time series at site 164 is highly structured and regular. This is consistent with its clear daily pattern, especially the repeated morning and evening traffic peaks.


```{r}
rqa_result <- crqa(
  ts1         = x_scaled,
  ts2         = x_scaled,
  delay       = 29,
  embed       = 5,
  rescale     = 0,
  radius      = 0.55,
  normalize   = 0,
  mindiagline = 2,
  minvertline = 2,
  tw          = 29,
  whiteline   = FALSE,
  recpt       = FALSE,
  side        = "both"
)

cat("RR:  ", rqa_result$RR,  "\n")
cat("DET: ", rqa_result$DET, "\n")
cat("L:   ", rqa_result$L,   "\n")
cat("LAM: ", rqa_result$LAM, "\n")
cat("TT:  ", rqa_result$TT,  "\n")
cat("ENTR:", rqa_result$ENTR,"\n")
```

```{r}
RP <- matrix(as.numeric(rqa_result$RP), nrow = nrow(rqa_result$RP))

plot(NULL, xlim = c(1, ncol(RP)), ylim = c(1, nrow(RP)),
     xlab = "Time (15-min intervals)",
     ylab = "Time (15-min intervals)",
     main = "Recurrence Plot - Site 164")

for (l in 1:ncol(RP)) {
  ind <- which(RP[, l] == 1)
  points(rep(l, length(ind)), ind,
         cex = 0.3,
         col = "blue4",
         pch = 16)
}
```

Prior to conducting the Recurrence Quantification Analysis (RQA), the time series of site 164 was standardized using z-score scaling. Consequently, the normalization and rescaling parameters within the crqa() function were both set to 0 (normalize = 0, rescale = 0), as the data had already been placed on a comparable scale. 

For the final RQA, the previously obtained parameters were used: τ = 29, M = 5, and radius = 0.55. The minimum diagonal line length and minimum vertical line length were both set to 2, as this is the minimum threshold required for a recurrent structure to be considered meaningful. The Theiler window was set to 29, matching the selected delay, to exclude points that were too close together in time and prevent trivial autocorrelation from being interpreted as meaningful recurrence. The analysis was conducted with side = "both", meaning that recurrence structures on both sides of the main diagonal were included.

The RQA results indicated a recurrence rate of 1.79%, determinism of 85.89%, a mean diagonal line length of 5.55, laminarity of 92.22%, a trapping time of 4.15, and entropy of 2.09. The recurrence rate was slightly lower than the 2.02% obtained during radius selection, which is expected because the Theiler window (tw = 29) was applied in the final analysis to exclude recurrent points located within a short temporal interval of one another near the line of identity, thereby removing trivial autocorrelated recurrences from the calculation.

The high determinism value (DET = 85.89%) indicates that most recurrent points formed diagonal line structures, suggesting that traffic states often reoccurred in structured temporal sequences. The mean diagonal line length (L = 5.55) suggests that these recurring sequences persisted for an average of approximately 83 minutes (5.55 × 15 minutes) before diverging. Laminarity was notably high (LAM = 92.22%), even higher than the determinism value, indicating that a large proportion of recurrent points formed vertical structures. This suggests that the system often remained in stable or slowly changing traffic states for consecutive observations. This is further supported by the trapping time (TT = 4.15), corresponding to approximately 62 minutes of sustained state stability on average. The entropy value (ENTR = 2.09) indicates some variation in the lengths of recurring diagonal structures, suggesting that although traffic patterns were highly structured, they were not completely uniform. Taken together, these findings suggest that traffic at site 164 showed strong temporal structure and stability, consistent with repeated daily traffic rhythms rather than random fluctuations.


## mdRQA - Multivariate Traffic System Analysis

```{r}
# Extract each site
site100 <- traffic_data[traffic_data$site_no == 100, ]
site161 <- traffic_data[traffic_data$site_no == 161, ]
site164 <- traffic_data[traffic_data$site_no == 164, ]

# Keep only timestamp and volume, and rename volume columns
site100 <- site100[, c("timestamp", "site_total_volume")]
site161 <- site161[, c("timestamp", "site_total_volume")]
site164 <- site164[, c("timestamp", "site_total_volume")]

colnames(site100)[2] <- "x100"
colnames(site161)[2] <- "x161"
colnames(site164)[2] <- "x164"

# Merge by shared timestamps only
traffic_joined <- merge(site100, site161, by = "timestamp")
traffic_joined <- merge(traffic_joined, site164, by = "timestamp")

# Order by timestamp
traffic_joined <- traffic_joined[order(traffic_joined$timestamp), ]

# Scale after alignment
traffic_matrix <- cbind(
  x100 = scale(traffic_joined$x100)[, 1],
  x161 = scale(traffic_joined$x161)[, 1],
  x164 = scale(traffic_joined$x164)[, 1]
)

# 2688 / 96 = 28 days
dim(traffic_matrix)
```

```{r, fig.height=2, fig.width=8}
ggplot(traffic_joined, aes(x = timestamp)) +
  geom_line(aes(y = x100, color = "Site 100")) +
  geom_line(aes(y = x161, color = "Site 161")) +
  geom_line(aes(y = x164, color = "Site 164")) +
  labs(y = "Traffic volume", color = "Site")
```

```{r, fig.height=2, fig.width=8}
ggplot(traffic_joined, aes(x = timestamp)) +
  geom_line(aes(y = scale(x100)[,1], color = "Site 100")) +
  geom_line(aes(y = scale(x161)[,1], color = "Site 161")) +
  geom_line(aes(y = scale(x164)[,1], color = "Site 164")) +
  labs(y = "Scale", color = "Site")
```

```{r}
# Convert to dataframe like they did
traffic_df <- as.data.frame(traffic_matrix)

nbins <- 10
# one full day
maxlag <- 96
criterion <- "firstBelow"
threshold <- exp(-1)

md_delay <- mdDelay(traffic_df, nbins, maxlag, criterion, threshold)
print(md_delay)
```

```{r}
tau <- md_delay
maxEmb <- 10
numSamples <- 500
Rtol <- 10
Atol <- 2

md_fnn <- mdFnn(traffic_df, tau, maxEmb, numSamples, Rtol, Atol)
plot(md_fnn$embTimes[1:length(md_fnn$fnnPerc)], md_fnn$fnnPerc, 
     type = 'l',
     xlab = "Embedding Dimension",
     ylab = "% False Nearest Neighbors",
     main = "mdFNN - All Three Sites")
print(md_fnn)

md_embed <- 2

```

```{r}
par <- list(method = "mdcrqa", metric = "euclidean", maxlag = 96,
            radiusspan = 100, radiussample = 40, normalize = 0,
            rescale = 0, mindiagline = 2, minvertline = 2, tw = 0,
            whiteline = FALSE, recpt = FALSE, side = "both",
            datatype = "continuous", fnnpercent = NA,
            typeami = NA, nbins = 50, criterion = "firstBelow",
            threshold = 1, maxEmb = 20, numSamples = 500,
            Rtol = 10, Atol = 2)

results <- optimizeParam(traffic_df, traffic_df, par, min.rec = 2, max.rec = 5)
print(unlist(results))
```

```{r}
delay <- md_delay
embed <- md_embed
rescale <- 0
radius <- results$radius
normalize <- 0
mindiagline <- 2
minvertline <- 2
tw <- 0
whiteline <- FALSE
recpt <- FALSE
side <- "both"
method <- "mdcrqa"
metric <- "euclidean"
datatype <- "continuous"

mdrqa_result <- crqa(traffic_df, traffic_df, delay, embed, rescale, radius,
                     normalize, mindiagline, minvertline, tw,
                     whiteline, recpt, side, method, metric, datatype)

cat("RR:  ", mdrqa_result$RR,  "\n")
cat("DET: ", mdrqa_result$DET, "\n")
cat("L:   ", mdrqa_result$L,   "\n")
cat("LAM: ", mdrqa_result$LAM, "\n")
cat("TT:  ", mdrqa_result$TT,  "\n")
cat("ENTR:", mdrqa_result$ENTR,"\n")

```

```{r}
RP <- matrix(as.numeric(mdrqa_result$RP), nrow = nrow(mdrqa_result$RP))

plot(NULL, xlim = c(1, ncol(RP)), ylim = c(1, nrow(RP)),
     xlab = "Time (15-min intervals)",
     ylab = "Time (15-min intervals)",
     main = "Recurrence Plot - All Sites (mdRQA)")

for (l in 1:ncol(RP)) {
  ind <- which(RP[, l] == 1)
  points(rep(l, length(ind)), ind,
         cex = 0.3, col = "blue4", pch = 16)
}
```

```{r}
res <- drpfromts(
  traffic_df[, 1],
  traffic_df[, 3],
  windowsize = 500,
  radius = results$radius,
  delay = md_delay,
  embed = md_embed,
  rescale = 0,
  normalize = 0,
  mindiagline = 2,
  minvertline = 2,
  tw = 0,
  whiteline = FALSE,
  recpt = FALSE,
  side = "both",
  method = "crqa",
  metric = "euclidean",
  datatype = "continuous"
)

profile <- res$profile

plot(seq(1, length(profile), 1), profile,
     type = "l", lwd = 2,
     xlab = "Lag",
     ylab = "Recurrence Rate",
     main = "Diagonal Cross-Recurrence Profile - Site 100 vs Site 164")
```
TODO: Improve DCRP plot
TODO: REPORT THE RESULTS OF mdRQA
TODO: Justify the selected parameters (Similar to RQA)


<!-- Try executing this chunk by clicking the *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. -->

<!-- Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*. -->

<!-- When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). -->

<!-- The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed. -->
